\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-04-24 doc/publications/plos16/nm.tex
add robs changes from plos16.tex to nm.tex

However, within an RDD, Spark ML stores each sample as a vector. Unlike RDDs, which can be partitioned and distributed
across multiple nodes, each vector must be present in its entirety on any node accessing it. This is no problem with
typical datasets, however as dimensionality increases, the vectors eventually reach a size where they can no longer fit
into a single node's memory.

@@

\subsection{Parameter settings}
We consider the parameter settings for the random forest algorithm. We use the R notation from the \randomforest
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
  importance;
\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry may give
  better performance;
\item the size and complexity of the individual trees is controlled in \randomforest by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest.\footnote{The Spark ML documentation \cite{Spark.2016} sets \texttt{maxDepth} to 4 in their
    classification example. This may give an estimate with a higher bias and, as such, may be a poor choice. See the
    discussion in \cite{Dietterich.2002} where it is shown that bagging may reduce bias as well as variance but give a
    better result for low bias learners. \cite{Dietterich.2002} notes that if the bootstrap replicate approximation were
    correct (i.e.~if the bootstrap sample came from an identical distribution to the data), then bagging would reduce
    variance without changing bias.}
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest can evaluate variable importance and perform classification for millions of features}
%\label{section:synthetic}
In this section we test the limits of \cursedforest\'s association testing and classify ability on high-dimensional data. 
%In this section we demonstrate the two features of \cursedforest, namely: variable selection analysis to select the most
%associated set of features amongst millions of dimensions; and its ability to accurately classify the labels of
%high-dimensional samples.  
We therefore generate a synthetic dataset with 2.5 million variables ($p=2500000$), of which 5 
are related to the response variable, and 5000 samples ($n=5000$), as discussed in the methods section.

>>>>

However, within an RDD, Spark ML stores each sample as a vector. Unlike RDDs, which can be partitioned and distributed
across multiple nodes, each vector must be present in its entirety on any node accessing it. This is no problem with
typical datasets; however as dimensionality increases, the vectors eventually reach a size where they can no longer fit
into a single node's memory.

@@

\subsection{Parameter settings}
We consider the parameter settings for the random forest algorithm. We use the R notation from the \randomforest
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
  importance;
\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry may give
  better performance;
\item the size and complexity of the individual trees is controlled in \randomforest by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest.\footnote{The Spark ML documentation \cite{Spark.2016} sets \texttt{maxDepth} to 4 in their
    classification example. This may give an estimate with a higher bias and, as such, may be a poor choice. See the
    discussion in \cite{Dietterich.2002} where it is shown that bagging may reduce bias as well as variance but is more
    complicated than first thought. They show that bagging may reduce bias as well as variance but give a    
    better result for low bias learners. \cite{Dietterich.2002} notes that if the bootstrap replicate approximation were
    correct (i.e.~if the bootstrap sample came from an identical distribution to the data), then bagging would reduce
    variance without changing bias.}
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest can evaluate variable importance and perform classification for millions of features}
%\label{section:synthetic}
In this section we test the limits of \cursedforest\'s association testing and classify ability on high-dimensional data. 
%In this section we demonstrate the two features of \cursedforest, namely: variable selection analysis to select the most
%associated set of features amongst millions of dimensions; and its ability to accurately classify the labels of
%high-dimensional samples.  
We therefore generate a synthetic dataset with 2.5 million variables ($p=2,500,000$), of which 5 
are related to the response variable, and 5000 samples ($n=5000$), as discussed in the methods section.
